# Introduction {#sec-intro}

This is where you introduce the main ideas of your thesis and provide an
overview of the context and background.

An honours thesis typically contains several chapters. This template
provides two example chapters — an introduction and a literature review — but
you should add or rename chapters to suit your research.

## Research question

State your research question clearly here.

## Thesis structure

Provide a brief overview of each chapter:

- **Chapter 2** reviews the relevant literature.
- **Chapter 3** describes the data and methodology.
- **Chapter 4** presents the results.
- **Chapter 5** concludes with a discussion and directions for future research.

## Using Quarto

The big advantage of Quarto is that it lets you include R or Python code
directly in your thesis, ensuring everything is reproducible and there are no
copy-paste errors.

For details on using Quarto, see <https://quarto.org>.

### Figures

```{r}
#| label: fig-example
#| fig-cap: "A simple example figure."
#| fig-width: 6
#| fig-height: 4
library(ggplot2)
mtcars |>
  ggplot(aes(x = wt, y = mpg)) +
  geom_point() +
  labs(x = "Weight (wt)", y = "Miles per gallon (mpg)")
```

@fig-example shows that heavier cars are less fuel-efficient, although EVs are very heavy and use no petrol at all.

### Tables

```{r}
#| label: tbl-example
#| tbl-cap: "A simple example table."
knitr::kable(
  data.frame(
    Variable = c("x", "y", "z"),
    Mean     = c(1.23, 4.56, 7.89),
    SD       = c(0.12, 0.45, 0.78)
  )
)
```

@tbl-example shows a simple summary table.

### Mathematics

Equations are typeset with full LaTeX-style support. For example, the normal
distribution density function is
$$
  f(x) = \frac{1}{\sigma\sqrt{2\pi}}
         \exp\left(-\frac{(x-\mu)^2}{2\sigma^2}\right).
$$ {#eq-normal}

### Cross-references

You can cross-reference sections (@sec-intro), figures (@fig-example),
tables (@tbl-example), and equations (@eq-normal).

### Citations

Citations use the `@key` syntax. For example, @Brown59 and @Winters60 are
references from the bibliography.
